---
title: "Analysis of the London Fire Brigade datasets:"
subtitle: "An investigation of the Incident Groups and Mobilisation records"
author: "Franklin Schram"
date: "2023-04-06"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Preamble

This dataset

## 2. Required libraries

You can also embed plots, for example:

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)      #
library(tidyr)      #
library(readr)      #
library(ggplot2)    #
library(scales)     # 
library(kableExtra) #
```

## 3. Downloading the datasets

### a. Prepping folders

```{bash folder preparation, message=FALSE, warning=FALSE}
DIR=/tmp/LFB_ANALYSIS
if [ -d "$DIR" ];
then
    echo "$DIR directory exists."
else
	echo "Creating workspace at /tmp/LFB_ANALYSIS"
	mkdir "$DIR directory does not exist."
fi

```

The data is available on the London Data Store

### b. Downloading the data

```{bash downloading the data, message=FALSE, warning=FALSE}

```

## 4. Reading in CSV and inspecting the data

```{r reading csv, message=FALSE, warning=FALSE, highlight=TRUE}
lfb_incidents = read_csv("LFB_Incidents.csv", col_names = TRUE, num_threads = 15)
```

```{r column names}
spec(lfb_incidents)
```

## 5. Investigating the incident file

We first start to investigate

### a. Overview of total incidents by group

```{r total incidents by group 2009-2023, message=FALSE, warning=FALSE, highlight=TRUE}
lfb_incidents_all <- lfb_incidents %>%
  select(IncidentGroup) %>%
  group_by(IncidentGroup) %>%
  summarise(total_count=n()) %>%
  arrange(total_count)
```

```{r total incidents by group 2009-2023 table, message=FALSE, warning=FALSE, highlight=TRUE}
kbl(lfb_incidents_all) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive",full_width = F))
```

```{r plot, message=FALSE, warning=FALSE, highlight=TRUE}
# Prevent scientific notation in charts
options(scipen = 999)

# Plotting with ggplot2
ggplot(data=lfb_incidents_all, aes(x=reorder(IncidentGroup, -total_count),
  y=total_count)) +
  coord_flip() +
  labs(
    title = "London Fire Brigade: Cumulative incidents recorded 2009-2023", 
    subtitle = "Broken down by Incident Groups",
    caption = "(Based on data from the London Fire Brigade)",
    x = "Incident Group", y = "Numbers of incidents recorded",) +
  geom_col()
```

```{r incidents per caldendar year, message=FALSE, warning=FALSE, highlight=TRUE}
lfb_incidents_per_cy <- lfb_incidents %>%
  select(IncidentGroup,CalYear) %>%
  group_by(IncidentGroup,CalYear) %>%
  summarise(total_count = n(), .groups = 'drop') %>%
  arrange(IncidentGroup, CalYear, total_count)
#  pivot_wider(
#    names_from = CalYear,
#    values_from = total_count
#  )
```

```{r plot incidents per calendar year, message=FALSE, warning=FALSE, highlight=TRUE}
ggplot(lfb_incidents_per_cy, aes(x = reorder(CalYear, -CalYear),
                                 y = total_count, 
                                 fill = reorder(IncidentGroup, -total_count))) +
  coord_flip() +
  labs(
    title = "London Fire Brigade: Incidents recorded per calendar year 2009-2023", 
    subtitle = "Broken down by Incident Groups",
    caption = "(Based on data from the London Fire Brigade)",
    x = "Incident Group", y = "Numbers recorded",) +
  geom_col(position = "dodge")
```

### b. Overview of false alarms

### c. Overview of fire incidents

```{r Fires breakdown, message=FALSE, warning=FALSE, highlight=TRUE}

lfb_incidents_fires <- lfb_incidents %>%
  select(StopCodeDescription, PropertyCategory) %>%
  filter(StopCodeDescription == "Primary Fire") %>%
  group_by(PropertyCategory) %>%
  summarise(total_count=n(), .groups = 'drop') %>%
  mutate(freq = round(total_count / sum(total_count) *100, 3)) %>%
  arrange(desc(total_count))
lfb_incidents_fires
```

### d. Overview of special services

```{r Special services breakdown, message=FALSE, warning=FALSE, highlight=TRUE}

lfb_incidents_ss <- lfb_incidents %>%
  select(SpecialServiceType) %>%
  filter( !is.na(SpecialServiceType)) %>%
  group_by(SpecialServiceType) %>%
  summarise(total_count=n()) %>%
  mutate(freq = round(total_count / sum(total_count) *100, 2)) %>%
  arrange(desc(total_count))
```

```{r special services breakdown table, message=FALSE, warning=FALSE, highlight=TRUE}
kbl(lfb_incidents_ss) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
  
```

```{r plottt, message=FALSE, warning=FALSE, highlight=TRUE}

# Plotting with ggplot2
ggplot(data=lfb_incidents_ss, aes(x=reorder(SpecialServiceType, -total_count),
  y=total_count)) +
  coord_flip() +
  labs(
    title = "London Fire Brigade: Special Services recorded 2009-2023", 
    subtitle = "Broken down by Type",
    caption = "(Data: London Fire Brigade)",
    x = "Special Service Type", y = "Numbers of incidents recorded",) +
  geom_text(aes(label = freq), hjust = -0.10, colour = "black") +
  geom_col()
```
